\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2025-12-09 misc/JSS article/final/article.tex
Replace sebkrantz.github.io/collapse -> fastverse.github.io/collapse.

Indexed by:  c.s [2] | y [9 (10)]
\end{Soutput}
\end{Schunk}
%
Last but not least, the computational performance of these classes is second to none.\footnote{See, e.g., the small benchmark presented on slide 40 of the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}.}
%
\section{Table joins and pivots} \label{sec:join_pivot}
%
Among its suite of \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions}, \pkg{collapse}'s implementations of table \href{https://sebkrantz.github.io/collapse/reference/join.html}{joins} and \href{https://sebkrantz.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and recast pivots. Both implementations also offer outstanding computational performance, syntax, and memory efficiency.
%
\subsection{Joins}
%
Compared to commercial software such as \proglang{STATA} \citep{STATA}, the implementation of joins in most open-source software provides no information on how many records were joined from both tables. This often provokes manual efforts to validate the join operation. \fct{collapse::join} provides many options to understand table join operations. Its syntax is:
\begin{Code}
join(x, y, on = NULL, how = "left", suffix = NULL, validate = "m:m",
  multiple = FALSE, sort = FALSE, keep.col.order = TRUE, verbose = 1,
  drop.dup.cols = FALSE, require = NULL, column = NULL, attr = NULL, ...)
\end{Code}
It defaults to left join and only takes first matches from \code{y} (\code{multiple = FALSE}). Thus, it simply adds columns to \code{x}, which is efficient and sufficient/desired in many cases. By default (\code{verbose = 1}), it prints information about the join operation and number of records joined. \newline

@@

\fct{collapse::pivot} provides a powerful new implementation of reshaping for \proglang{R} addressing these shortcomings. It has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots and supports complex labelled data without loss of information. Its basic syntax is: \newpage
%
\begin{Code}
pivot(data, ids = NULL, values = NULL, names = NULL, labels = NULL,
  how = "longer", na.rm = FALSE, check.dups = FALSE, FUN = "last", ...)
\end{Code}
%
The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Database (\href{https://sebkrantz.github.io/collapse/reference/GGDC10S.html}{\code{GGDC10S}}) providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America. While the database covers 10 sectors, for the demonstration I only retain Agriculture, Mining, and Manufacturing.\footnote{The \code{"Label"} column is added for demonstration purposes. \fct{namlab} provides a compact overview of variable names and labels stored in \code{attr(column, "label")}, with (optional) additional information/statistics.}
%
\begin{Schunk}
\begin{Sinput}
R> data <- GGDC10S |>
+   fmutate(Label = ifelse(Variable == "VA", "Value Added", "Employment")) |>
+   fsubset(is.finite(AGR), Country, Variable, Label, Year, AGR:MAN)
R> namlab(data, N = TRUE, Ndistinct = TRUE, class = TRUE)
\end{Sinput}
\begin{Soutput}
  Variable     Class    N Ndist         Label
1  Country character 4364    43       Country
2 Variable character 4364     2      Variable
3    Label character 4364     2          <NA>
4     Year   numeric 4364    67          Year
5      AGR   numeric 4364  4353   Agriculture
6      MIN   numeric 4355  4224        Mining
7      MAN   numeric 4355  4353 Manufacturing
\end{Soutput}
\end{Schunk}
%
To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = 1:4)}. If \code{labels = "name"} is specified, variable labels stored in \code{attr(column, "label")} are saved to an additional column. In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to set alternative names for the \code{variable} and \code{value} columns.
%
\begin{Schunk}
\begin{Sinput}
R> head(dl <- pivot(data, ids = 1:4, names = list("Sectorcode", "Value"),
+    labels = "Sector", how = "longer"))
\end{Sinput}
\begin{Soutput}
  Country Variable       Label Year Sectorcode      Sector Value
1     BWA       VA Value Added 1964        AGR Agriculture 16.30
2     BWA       VA Value Added 1965        AGR Agriculture 15.73
3     BWA       VA Value Added 1966        AGR Agriculture 17.68
4     BWA       VA Value Added 1967        AGR Agriculture 19.15
5     BWA       VA Value Added 1968        AGR Agriculture 21.10
6     BWA       VA Value Added 1969        AGR Agriculture 21.86
\end{Soutput}
\end{Schunk}
%
\fct{pivot} only requires essential information and intelligently guesses the rest. For example, the same result could have been obtained by specifying \code{values = c("AGR", "MIN", "MAN")} instead of \code{ids = 1:4}. An exact reverse operation can also be specified as \code{pivot(dl, 1:4, "Value", "Sectorcode", "Sector", how = "wider")}, where \code{dl} is the long data. \newline

@@

\fct{pivot} also supports fast aggregation pivots, the default being \code{FUN = "last"}, which simply overwrites values in appearance order if the combination of \code{ids} and \code{names} does not fully identify the data. The latter can be checked with \code{check.dups = TRUE}. A small number extremely fast internal aggregation functions: \code{"first", "last", "sum", "mean", "min", "max",} and \code{"count"}, operate 'on the fly' during reshaping. \fct{pivot} also supports \emph{Fast Statistical Functions}, which will yield vectorized aggregations, but requires a deep copy of the columns aggregated which is avoided by the internal functions. The following example performs aggregation across years with the internal mean function during a recast pivot.
\begin{Schunk}
\begin{Sinput}
R> head(dr_agg <- pivot(data, "Country", c("AGR", "MIN", "MAN"),
+    how = "recast", names = list(from = "Variable", to = "Sectorcode"),
+    labels = list(from = "Label", to = "Sector"), FUN = "mean"))
\end{Sinput}
\begin{Soutput}
  Country Sectorcode      Sector       VA      EMP
1     BWA        AGR Agriculture    462.2   188.06
2     ETH        AGR Agriculture  34389.9 17624.34
3     GHA        AGR Agriculture   1549.4  3016.04
4     KEN        AGR Agriculture 139705.9  5348.91
5     MWI        AGR Agriculture  28512.6  2762.62
6     MUS        AGR Agriculture   3819.6    59.34
\end{Soutput}
\end{Schunk}
The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples} demonstrate more features of \fct{pivot}. Notably, it can also perform longer and recast pivots without id variables, similar to \code{data.table::transpose()}.
%
\section{List processing} \label{sec:list_proc}
%
Often in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects, such as a model predictions and performance statistics, in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices. Except for certain recursive functions in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit. \newline

@@

\pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious yet powerful toolkit to examine complex datasets. A particular focus has been on providing tools to understand multilevel (panel) data. Recall the World Development panel dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) from Section~\ref{sec:integration}. The function \fct{varying} can be used to examine which of the variables are time-varying.

@@

%
\fct{descr} also has a \code{stepwise} argument to describe one variable at a time, allowing users to naturally 'step-through' the variables in a large dataset while spreading the computational burden. The \href{https://sebkrantz.github.io/collapse/reference/descr.html}{documentation} provides more details and examples. Both \fct{qsu} and \fct{descr} come with an \code{as.data.frame()} method for efficient tidying and easy further analysis. \newline

A final noteworthy function from \pkg{collapse}'s descriptive statistics toolkit is \fct{qtab}, an enhanced drop-in replacement for \fct{table}. It is enhanced both in a statistical and a computational sense, providing a remarkable performance boost, an option \code{sort = FALSE} to preserve the first-appearance-order of vectors being cross-tabulated, support for frequency weights (\code{w}), and the ability to compute different statistics representing table entries using these weights---vectorized when using \emph{Fast Statistical Functions} as demonstrated below.
%
\begin{Schunk}
\begin{Sinput}
R> wlda15 |> with(qtab(OECD, income))
\end{Sinput}
\begin{Soutput}
       income
OECD    High income Low income Lower middle income Upper middle income
  FALSE          45         30                  47                  58
  TRUE           34          0                   0                   2
\end{Soutput}
\end{Schunk}
This shows the total population (latest post-2015 estimates) in millions.
\begin{Schunk}
\begin{Sinput}
R> wlda15 |> with(qtab(OECD, income, w = POP / 1e6))
\end{Sinput}
\begin{Soutput}
       income
OECD    High income Low income Lower middle income Upper middle income
  FALSE       93.01     694.89             3063.54             2459.71
  TRUE      1098.75       0.00                0.00              211.01
\end{Soutput}
\end{Schunk}
This shows the average life expectancy in years. The use of \fct{fmean} toggles an efficient vectorized computation of the table entries (i.e., \fct{fmean} is only called once).
\begin{Schunk}
\begin{Sinput}
R> wlda15 |> with(qtab(OECD, income, w = LIFEEX, wFUN = fmean))
\end{Sinput}
\begin{Soutput}
       income
OECD    High income Low income Lower middle income Upper middle income
  FALSE       78.75      62.81               68.30               73.81
  TRUE        81.09                                              76.37
\end{Soutput}
\end{Schunk}
Finally, this calculates a population-weighted average of life expectancy in each group.
\begin{Schunk}
\begin{Sinput}
R> wlda15 |> with(qtab(OECD, income, w = LIFEEX, wFUN = fmean,
+                      wFUN.args = list(w = POP)))
\end{Sinput}
\begin{Soutput}
       income
OECD    High income Low income Lower middle income Upper middle income
  FALSE       77.91      63.81               68.76               75.93
  TRUE        81.13                                              76.10
\end{Soutput}
\end{Schunk}
%
\class{qtab} objects inherit the \class{table} class, thus all \class{table} methods apply. Apart from the above summary functions, \pkg{collapse} also provides \fct{pwcor}, \fct{pwcov}, and \fct{pwnobs} for convenient (pairwise, weighted) correlations, covariances, and observations counts, respectively.
%
\section{Global options} \label{sec:glob_opt}
%
\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/reference/collapse-options.html}{globally configurable} to an extent few packages are: the default values of key function arguments governing the behavior of its algorithms, and the exported namespace, can be adjusted interactively through the \fct{set\_collapse} function. These options are saved in an internal environment called \code{.op}. Its contents can be accessed using \fct{get\_collapse}. \newline

@@

To reduce this cost, \pkg{collapse}'s \code{mask} option allows masking existing \proglang{R} functions with the faster \pkg{collapse} versions by creating additional functions in the namespace and instantly exporting them. All \pkg{collapse} functions starting with an 'f' can be passed to the option (with or without the 'f'), e.g., \code{set\_collapse(mask = c("subset", "transform"))} creates \code{subset <- fsubset} and \code{transform <- ftransform} and exports them. Special functions are \code{"n"}, \code{"table"/"qtab"}, and \code{"\%in\%"}, which create \code{n <- GRPN} for use in \code{(f)summarise}/\code{(f)mutate}, \code{table <- qtab}, and replace \code{\%in\%} with a faster version based on \fct{fmatch}, respectively. There also exist several \href{https://sebkrantz.github.io/collapse/reference/collapse-options.html}{convenience keywords to mask related groups of functions}, such as \code{"manip"} (only data manipulation functions), or \code{"all"} (everything), as demonstrated below.
%
\begin{Code}
set_collapse(mask = "all")
wlddev |> subset(year >= 1990 & is.finite(GINI)) |>
  group_by(year) |>
  summarise(n = n(), across(PCGDP:GINI, mean, w = POP))
with(mtcars, table(cyl, vs, am))
sum(mtcars)
diff(EuStockMarkets)
mean(num_vars(iris), g = iris$Species)
unique(wlddev, cols = c("iso3c", "year"))
range(wlddev$date)
wlddev |> index_by(iso3c, year) |>
  mutate(PCGDP_lag = lag(PCGDP),
         PCGDP_growth = growth(PCGDP)) |> unindex()
\end{Code}
%
The above is now 100\% \pkg{collapse} code. Similarly, using this option, all code in this article could have been written without f-prefixes. Thus, \pkg{collapse} is able to offer a fast and syntactically clean experience of \proglang{R}---without the need to even restart the session. Masking is completely and interactively reversible: calling \code{set\_collapse(mask = NULL)} instantly removes the additional functions. Option \code{remove} can further be used to remove (un-export) any \pkg{collapse} function, allowing manual conflict management. Function \code{fastverse::fastverse\_conflicts()} from the related \href{https://fastverse.github.io/fastverse/}{\pkg{fastverse} project} \citep{rfastverse} can be used to display namespace conflicts with \pkg{collapse}. Invoking either \code{mask} or \code{remove} detaches \pkg{collapse} and re-attaches it at the top of the search path, letting its namespace to take precedence over other packages. \newline

Such global powers confer responsibilities upon package developers, as further elucidated in the \href{https://sebkrantz.github.io/collapse/articles/developing_with_collapse.html#some-notes-on-global-options}{vignette on developing with \pkg{collapse}}. As a general rule, options \code{mask} and \code{remove} should be off-limits inside packages, and other options need to be reset immediately using \fct{on.exit}.

>>>>

Indexed by:  c.s [2] | y [9 (10)]
\end{Soutput}
\end{Schunk}
%
Last but not least, the computational performance of these classes is second to none.\footnote{See, e.g., the small benchmark presented on slide 40 of the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}.}
%
\section{Table joins and pivots} \label{sec:join_pivot}
%
Among its suite of \href{https://fastverse.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions}, \pkg{collapse}'s implementations of table \href{https://fastverse.github.io/collapse/reference/join.html}{joins} and \href{https://fastverse.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and recast pivots. Both implementations also offer outstanding computational performance, syntax, and memory efficiency.
%
\subsection{Joins}
%
Compared to commercial software such as \proglang{STATA} \citep{STATA}, the implementation of joins in most open-source software provides no information on how many records were joined from both tables. This often provokes manual efforts to validate the join operation. \fct{collapse::join} provides many options to understand table join operations. Its syntax is:
\begin{Code}
join(x, y, on = NULL, how = "left", suffix = NULL, validate = "m:m",
  multiple = FALSE, sort = FALSE, keep.col.order = TRUE, verbose = 1,
  drop.dup.cols = FALSE, require = NULL, column = NULL, attr = NULL, ...)
\end{Code}
It defaults to left join and only takes first matches from \code{y} (\code{multiple = FALSE}). Thus, it simply adds columns to \code{x}, which is efficient and sufficient/desired in many cases. By default (\code{verbose = 1}), it prints information about the join operation and number of records joined. \newline

@@

\fct{collapse::pivot} provides a powerful new implementation of reshaping for \proglang{R} addressing these shortcomings. It has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots and supports complex labelled data without loss of information. Its basic syntax is: \newpage
%
\begin{Code}
pivot(data, ids = NULL, values = NULL, names = NULL, labels = NULL,
  how = "longer", na.rm = FALSE, check.dups = FALSE, FUN = "last", ...)
\end{Code}
%
The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Database (\href{https://fastverse.github.io/collapse/reference/GGDC10S.html}{\code{GGDC10S}}) providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America. While the database covers 10 sectors, for the demonstration I only retain Agriculture, Mining, and Manufacturing.\footnote{The \code{"Label"} column is added for demonstration purposes. \fct{namlab} provides a compact overview of variable names and labels stored in \code{attr(column, "label")}, with (optional) additional information/statistics.}
%
\begin{Schunk}
\begin{Sinput}
R> data <- GGDC10S |>
+   fmutate(Label = ifelse(Variable == "VA", "Value Added", "Employment")) |>
+   fsubset(is.finite(AGR), Country, Variable, Label, Year, AGR:MAN)
R> namlab(data, N = TRUE, Ndistinct = TRUE, class = TRUE)
\end{Sinput}
\begin{Soutput}
  Variable     Class    N Ndist         Label
1  Country character 4364    43       Country
2 Variable character 4364     2      Variable
3    Label character 4364     2          <NA>
4     Year   numeric 4364    67          Year
5      AGR   numeric 4364  4353   Agriculture
6      MIN   numeric 4355  4224        Mining
7      MAN   numeric 4355  4353 Manufacturing
\end{Soutput}
\end{Schunk}
%
To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = 1:4)}. If \code{labels = "name"} is specified, variable labels stored in \code{attr(column, "label")} are saved to an additional column. In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to set alternative names for the \code{variable} and \code{value} columns.
%
\begin{Schunk}
\begin{Sinput}
R> head(dl <- pivot(data, ids = 1:4, names = list("Sectorcode", "Value"),
+    labels = "Sector", how = "longer"))
\end{Sinput}
\begin{Soutput}
  Country Variable       Label Year Sectorcode      Sector Value
1     BWA       VA Value Added 1964        AGR Agriculture 16.30
2     BWA       VA Value Added 1965        AGR Agriculture 15.73
3     BWA       VA Value Added 1966        AGR Agriculture 17.68
4     BWA       VA Value Added 1967        AGR Agriculture 19.15
5     BWA       VA Value Added 1968        AGR Agriculture 21.10
6     BWA       VA Value Added 1969        AGR Agriculture 21.86
\end{Soutput}
\end{Schunk}
%
\fct{pivot} only requires essential information and intelligently guesses the rest. For example, the same result could have been obtained by specifying \code{values = c("AGR", "MIN", "MAN")} instead of \code{ids = 1:4}. An exact reverse operation can also be specified as \code{pivot(dl, 1:4, "Value", "Sectorcode", "Sector", how = "wider")}, where \code{dl} is the long data. \newline

@@

\fct{pivot} also supports fast aggregation pivots, the default being \code{FUN = "last"}, which simply overwrites values in appearance order if the combination of \code{ids} and \code{names} does not fully identify the data. The latter can be checked with \code{check.dups = TRUE}. A small number extremely fast internal aggregation functions: \code{"first", "last", "sum", "mean", "min", "max",} and \code{"count"}, operate 'on the fly' during reshaping. \fct{pivot} also supports \emph{Fast Statistical Functions}, which will yield vectorized aggregations, but requires a deep copy of the columns aggregated which is avoided by the internal functions. The following example performs aggregation across years with the internal mean function during a recast pivot.
\begin{Schunk}
\begin{Sinput}
R> head(dr_agg <- pivot(data, "Country", c("AGR", "MIN", "MAN"),
+    how = "recast", names = list(from = "Variable", to = "Sectorcode"),
+    labels = list(from = "Label", to = "Sector"), FUN = "mean"))
\end{Sinput}
\begin{Soutput}
  Country Sectorcode      Sector       VA      EMP
1     BWA        AGR Agriculture    462.2   188.06
2     ETH        AGR Agriculture  34389.9 17624.34
3     GHA        AGR Agriculture   1549.4  3016.04
4     KEN        AGR Agriculture 139705.9  5348.91
5     MWI        AGR Agriculture  28512.6  2762.62
6     MUS        AGR Agriculture   3819.6    59.34
\end{Soutput}
\end{Schunk}
The \href{https://fastverse.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples} demonstrate more features of \fct{pivot}. Notably, it can also perform longer and recast pivots without id variables, similar to \code{data.table::transpose()}.
%
\section{List processing} \label{sec:list_proc}
%
Often in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects, such as a model predictions and performance statistics, in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices. Except for certain recursive functions in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://fastverse.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit. \newline

@@

\pkg{collapse}'s \href{https://fastverse.github.io/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious yet powerful toolkit to examine complex datasets. A particular focus has been on providing tools to understand multilevel (panel) data. Recall the World Development panel dataset (\href{https://fastverse.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) from Section~\ref{sec:integration}. The function \fct{varying} can be used to examine which of the variables are time-varying.

@@

%
\fct{descr} also has a \code{stepwise} argument to describe one variable at a time, allowing users to naturally 'step-through' the variables in a large dataset while spreading the computational burden. The \href{https://fastverse.github.io/collapse/reference/descr.html}{documentation} provides more details and examples. Both \fct{qsu} and \fct{descr} come with an \code{as.data.frame()} method for efficient tidying and easy further analysis. \newline

A final noteworthy function from \pkg{collapse}'s descriptive statistics toolkit is \fct{qtab}, an enhanced drop-in replacement for \fct{table}. It is enhanced both in a statistical and a computational sense, providing a remarkable performance boost, an option \code{sort = FALSE} to preserve the first-appearance-order of vectors being cross-tabulated, support for frequency weights (\code{w}), and the ability to compute different statistics representing table entries using these weights---vectorized when using \emph{Fast Statistical Functions} as demonstrated below.
%
\begin{Schunk}
\begin{Sinput}
R> wlda15 |> with(qtab(OECD, income))
\end{Sinput}
\begin{Soutput}
       income
OECD    High income Low income Lower middle income Upper middle income
  FALSE          45         30                  47                  58
  TRUE           34          0                   0                   2
\end{Soutput}
\end{Schunk}
This shows the total population (latest post-2015 estimates) in millions.
\begin{Schunk}
\begin{Sinput}
R> wlda15 |> with(qtab(OECD, income, w = POP / 1e6))
\end{Sinput}
\begin{Soutput}
       income
OECD    High income Low income Lower middle income Upper middle income
  FALSE       93.01     694.89             3063.54             2459.71
  TRUE      1098.75       0.00                0.00              211.01
\end{Soutput}
\end{Schunk}
This shows the average life expectancy in years. The use of \fct{fmean} toggles an efficient vectorized computation of the table entries (i.e., \fct{fmean} is only called once).
\begin{Schunk}
\begin{Sinput}
R> wlda15 |> with(qtab(OECD, income, w = LIFEEX, wFUN = fmean))
\end{Sinput}
\begin{Soutput}
       income
OECD    High income Low income Lower middle income Upper middle income
  FALSE       78.75      62.81               68.30               73.81
  TRUE        81.09                                              76.37
\end{Soutput}
\end{Schunk}
Finally, this calculates a population-weighted average of life expectancy in each group.
\begin{Schunk}
\begin{Sinput}
R> wlda15 |> with(qtab(OECD, income, w = LIFEEX, wFUN = fmean,
+                      wFUN.args = list(w = POP)))
\end{Sinput}
\begin{Soutput}
       income
OECD    High income Low income Lower middle income Upper middle income
  FALSE       77.91      63.81               68.76               75.93
  TRUE        81.13                                              76.10
\end{Soutput}
\end{Schunk}
%
\class{qtab} objects inherit the \class{table} class, thus all \class{table} methods apply. Apart from the above summary functions, \pkg{collapse} also provides \fct{pwcor}, \fct{pwcov}, and \fct{pwnobs} for convenient (pairwise, weighted) correlations, covariances, and observations counts, respectively.
%
\section{Global options} \label{sec:glob_opt}
%
\pkg{collapse} is \href{https://fastverse.github.io/collapse/reference/collapse-options.html}{globally configurable} to an extent few packages are: the default values of key function arguments governing the behavior of its algorithms, and the exported namespace, can be adjusted interactively through the \fct{set\_collapse} function. These options are saved in an internal environment called \code{.op}. Its contents can be accessed using \fct{get\_collapse}. \newline

@@

To reduce this cost, \pkg{collapse}'s \code{mask} option allows masking existing \proglang{R} functions with the faster \pkg{collapse} versions by creating additional functions in the namespace and instantly exporting them. All \pkg{collapse} functions starting with an 'f' can be passed to the option (with or without the 'f'), e.g., \code{set\_collapse(mask = c("subset", "transform"))} creates \code{subset <- fsubset} and \code{transform <- ftransform} and exports them. Special functions are \code{"n"}, \code{"table"/"qtab"}, and \code{"\%in\%"}, which create \code{n <- GRPN} for use in \code{(f)summarise}/\code{(f)mutate}, \code{table <- qtab}, and replace \code{\%in\%} with a faster version based on \fct{fmatch}, respectively. There also exist several \href{https://fastverse.github.io/collapse/reference/collapse-options.html}{convenience keywords to mask related groups of functions}, such as \code{"manip"} (only data manipulation functions), or \code{"all"} (everything), as demonstrated below.
%
\begin{Code}
set_collapse(mask = "all")
wlddev |> subset(year >= 1990 & is.finite(GINI)) |>
  group_by(year) |>
  summarise(n = n(), across(PCGDP:GINI, mean, w = POP))
with(mtcars, table(cyl, vs, am))
sum(mtcars)
diff(EuStockMarkets)
mean(num_vars(iris), g = iris$Species)
unique(wlddev, cols = c("iso3c", "year"))
range(wlddev$date)
wlddev |> index_by(iso3c, year) |>
  mutate(PCGDP_lag = lag(PCGDP),
         PCGDP_growth = growth(PCGDP)) |> unindex()
\end{Code}
%
The above is now 100\% \pkg{collapse} code. Similarly, using this option, all code in this article could have been written without f-prefixes. Thus, \pkg{collapse} is able to offer a fast and syntactically clean experience of \proglang{R}---without the need to even restart the session. Masking is completely and interactively reversible: calling \code{set\_collapse(mask = NULL)} instantly removes the additional functions. Option \code{remove} can further be used to remove (un-export) any \pkg{collapse} function, allowing manual conflict management. Function \code{fastverse::fastverse\_conflicts()} from the related \href{https://fastverse.github.io/fastverse/}{\pkg{fastverse} project} \citep{rfastverse} can be used to display namespace conflicts with \pkg{collapse}. Invoking either \code{mask} or \code{remove} detaches \pkg{collapse} and re-attaches it at the top of the search path, letting its namespace to take precedence over other packages. \newline

Such global powers confer responsibilities upon package developers, as further elucidated in the \href{https://fastverse.github.io/collapse/articles/developing_with_collapse.html#some-notes-on-global-options}{vignette on developing with \pkg{collapse}}. As a general rule, options \code{mask} and \code{remove} should be off-limits inside packages, and other options need to be reset immediately using \fct{on.exit}.

### 2025-12-09 misc/JSS article/final/jss5278/jss5278.tex
Replace sebkrantz.github.io/collapse -> fastverse.github.io/collapse.

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://sebkrantz.github.io/collapse/. Linecount (v2.1.2): \proglang{R}: 13,785, \proglang{C}: 19,013, \proglang{C++}: 9,844. \hphantom{aaa.}Exported namespace: 392 objects, of which 238 functions (excl. methods and shorthands), and 2 datasets.}
% Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently.
Core functionality includes a rich set of S3 generic (grouped, weighted) statistical functions for vectors, matrices, and data frames, which provide efficient low-level vectorizations, OpenMP multithreading, and skip missing values by default (\code{na.rm = TRUE}). It also provides powerful data manipulation functions, including vectorized and verbose hash-joins and fast (aggregation, recast) pivots, functions and classes for indexed (time-aware) computations on time series and panel data, recursive tools to deal with nested data, and advanced descriptive statistical tools. This functionality is powered by efficient algorithms for grouping, ordering, deduplication, and matching callable at both \proglang{R} and \proglang{C} levels. The package also includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data. \pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames, and seamlessly supporting popular extensions to these objects such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}. It is also extensively globally and interactively configurable. \newline

@@

This article cannot fully present \pkg{collapse}, but the following sections introduce its key features, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options. Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes. For deeper engagement with \pkg{collapse}, consult the \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://sebkrantz.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk} with \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.

>>>>

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://fastverse.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://fastverse.github.io/collapse/. Linecount (v2.1.2): \proglang{R}: 13,785, \proglang{C}: 19,013, \proglang{C++}: 9,844. \hphantom{aaa.}Exported namespace: 392 objects, of which 238 functions (excl. methods and shorthands), and 2 datasets.}
% Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently.
Core functionality includes a rich set of S3 generic (grouped, weighted) statistical functions for vectors, matrices, and data frames, which provide efficient low-level vectorizations, OpenMP multithreading, and skip missing values by default (\code{na.rm = TRUE}). It also provides powerful data manipulation functions, including vectorized and verbose hash-joins and fast (aggregation, recast) pivots, functions and classes for indexed (time-aware) computations on time series and panel data, recursive tools to deal with nested data, and advanced descriptive statistical tools. This functionality is powered by efficient algorithms for grouping, ordering, deduplication, and matching callable at both \proglang{R} and \proglang{C} levels. The package also includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data. \pkg{collapse} is \href{https://fastverse.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames, and seamlessly supporting popular extensions to these objects such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}. It is also extensively globally and interactively configurable. \newline

@@

This article cannot fully present \pkg{collapse}, but the following sections introduce its key features, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options. Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes. For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://fastverse.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk} with \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
